---
title: 'Lab 3: Linear Regression'
author:
- Farwa Bhatti
- Deepa Borkar
- Justin Hsia
date: "November 20, 2019"
output:
  github_document: default
  pdf_document: default
subtitle: An analysis of dataset of crime statistics for a selection of counties in
  North Carolina
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup1, include=FALSE, echo=FALSE}
library(rmarkdown)
library(car)
library(stargazer)
library(lmtest)
library(Hmisc)
library(corrplot)
library(effsize)
```

## 1. Introduction

The team is tasked to carry out research for a political campaign through a dataset of crime statistics for a selection of counties in North Carolina.  Using statistical modeling approach, the team aims to explore the determinants of crime for the given selection of counties in North Carolina, and suggest applicable policies which will be helpful to the local government to reduce crime.  The following research question and the supplementary questions will guide the recommendations and lead to an actionable policy.

*Research Question*: What factors affect the crime rate for the given selection of counties in North Carolina?

*Supplementary research questions to support the modeling process*:

1. What factors , potentially controlled by the local government, affect the crime rate?

2. What factors, which are important but may not be controlled by the local government, affect the crime rate?

3. What is the magnitude effect of these factors, that is, are some factors more important than others in contributing to the crime rate?

4. Do these results vary by region, area,  type of criminal activity etc?

Throughout the rest of our report, our objective is to best answer the main research question and to provide as much statistical evidence and support for the supplementary research questions.  However given the scope of the project and the limitations of the dataset, we may only be able to answer some of those questions.  The end goal is to provide the policy makers with actionable insights that they will be able to implement to control the crime rate in North Carolina.

\  

## 2. Exploratory Data Analysis

**Variables**: The dataset has 25 variables, with ***crmrte - crimes committed per person*** being the dependent variable in our statistical models.  Before we start brainstorming on the modelling approaches, it is important to validate and explore the variables of interest from the dataset, and to detect and correct the possibly inaccurate, incomplete or irrelevant records from the dataset.

```{r, echo=TRUE}
C = read.csv('crime_v2.csv')

# Dimension of the database
dim(C)

# Objects of the databse
ls(C)
```  

\  
  
**Missing data**: The data initially had 97 rows and 25 columns. However, further looking into the data, we observed about 6 rows with no data labeled as “NA”, which also impacted some of the initial summary statistics.  The missing data was deleted from the final dataset which will be used for modelling purposes.  It is important to note that normally if we had some observations missing from the dataset, we would have used some extrapolation/imputation techniques for those missing values but given that we have 6 entire rows of NAs with no added information in them, we have decided that in order to improve the quality of the data and to derive meaningful insights from the EDA, it is okay to omit those rows.

```{r, echo=TRUE}
# Summary of the database
summary(C)

# First six sample points
head(C)

# Last six sample points
tail(C)

# Remove last six sample points for they hold no values
newC = C[1:91,]
```

\  

**Conversion of a variable**: It was also observed that the variable prbconv was a factor, and therefore the variable was converted from a factor to a numerical vector.

```{r, echo=TRUE}
# Check class of prbconv
class(newC$prbconv)

# Convert prbconv to a numerical vector
newC$prbconv = as.numeric(newC$prbconv)
```

\  

**Inaccurate observations**: There are also some probability variables which we further investigated to ensure that their values are in line with the definitions provided.

* ***prbarr - the probability of arrest*** - is the ratio of arrests to offenses. 

* ***prbconv - the probability of conviction*** - is the ratio of convictions to arrests

* ***prbpris - the probability of prison*** - ratio of convictions resulting in prison sentence to total convictions.

Given that these variables are probabilities and their values should reside between 0 and 1; we saw some observations where the values of these variables were greater than 1, and require deeper analysis as to how these values were computed.  One observation in ***prbarr*** and ten entries in ***prbconv*** have probabilities greater than 0.  We believe either there are mistakes in the database or there may be need to study these variables in greater detail.  Sticking to the definition of these variables as ***“probabilities”*** and limiting the scope of the project provided, any probability values greater than 1 can be interpreted as incorrect (or skewed) and will be removed from the dataset.

```{r, echo=TRUE}
length(newC$prbarr[newC$prbarr > 1])
length(newC$prbconv[newC$prbconv > 1])
length(newC$prbpris[newC$prbpris > 1])

# Filter out entries with probabilities that are greater than 1.
filter = newC$prbarr <= 1 & newC$prbconv <= 1 & newC$prbpris <= 1
filterC = newC[filter,]
```

\  

**Correlation Matrix**:  In order to understand the statistical associations of all of our variables, and which ones can potentially be the explanatory variables when we build our statistical models, we mapped out a correlation heat map matrix.  As highlighted in the **Exhibit EDA1: Correlation Heat Map Matrix** below, the red-to-orange toned variables are negatively correlated and the blue-to-dark blue toned variables are positively correlated.  The size and the color determine the magnitude of the correlation coefficient.

The two variables which in comparison to the other variables, have a higher correlation with the crime rate are *density* (population density) and *polpc* (police per capita).  The correlation coefficients are 0.72 and 0.48, respectively.  However, there are also some variables which are negatively correlated to *crmrte*, such as *prbconv* (probability of conviction) and *prbarr* (probability of arrest).  The negative correlation coefficients are not very strong, and are -0.3 and -0.5 for *prbconv* and *prbarr*, respectively.  Out of all the variables, population density has the strongest correlation and may also be one of the most important explanatory variables from the dataset.

```{r, echo=FALSE}
cmatrix = rcorr(as.matrix(filterC[, 3:25]))$r
corrplot(cmatrix, type="upper", order="hclust", tl.col="black", tl.srt=45)
```

\  

**Summary Statistics**: After the aforementioned changes, the filtered/cleaned dataset has a total of 81 rows and 25 variables, and all probability variables now have values less than or equal to 1.  For the rest of the analysis and the model building, we will use the cleaned dataset.

Our primary variable of interest given the research question is *crmrte*, which is crimes committed per person in each county.  The focus is via modeling approaches, we discover any causal or associative relations between the dependent variable *crmrte* and other variables. 

```{r new_summary, echo=FALSE}
summary(filterC)
```

\  

**EXPLORING VARIABLES WITH UNIVARIATE ANALYSIS- DENSITY**: For the univariate analysis, between population and crime rate, we observe that the model achieves R-squared of about 52% with a statistically significant coefficient.  One unit increase in population density increases the crime rate by 0.009 units.  It is also observed that the density histogram is positively skewed, however we may assume normality given the number of observations in the dataset for linear regression models.

```{r, echo=FALSE, fig.width=3.5, fig.height=3, fig.pos="H"}
hist(filterC$density, breaks = 20, main = "Histogram of Population Density", xlab = "Population Density")
model1 = lm(crmrte ~ density, data = filterC)
plot(filterC$density, filterC$crmrte, main = "Crime Rate vs Population Density", xlab = "Population Density", ylab = "Crime Rate")
abline(model1)
summary(model1)$r.square
stargazer(model1, type = "text",  out = "model1.txt")
```

\  

**EXPLORING VARIABLES WITH UNIVARIATE ANALYSIS- POLPC**: For the second univariate analysis, between police per capita and crime rate, we observe that the model achieves R-squared of about 31% with a statistically significant coefficient.  We also observe the magnitude of this coefficient to be high which can either over power the model results, or it is due to the difference in the per capita unit.  We may also assume normality in the histogram for *polpc*.

```{r, echo=FALSE, fig.width=3.5, fig.height=3, fig.pos="H"}
hist(filterC$polpc, breaks = 20, main = "Histogram of Police Per Capita", xlab = "Police Per Capita")
model2 = lm(crmrte ~ polpc, data = filterC)
plot(filterC$polpc, filterC$crmrte, main = "Crime Rate vs Police Per Capita", xlab = "Police Per Capita", ylab = "Crime Rate")
abline(model2)
summary(model2)$r.square
stargazer(model2, type = "text",  out = "model2.txt")
```

\  

**ISSUES WITH MULTICOLLINEARITY**: We ensure before we start the modelling approaches, that the variables we choose, (*density, polpc, ymlepsq*) are not correlated to one another.  We will define variable, **ymlepsq**, in the next section.

\  

**MODEL1 WITH EXPLANATORY VARIABLE:**

The first model finalized uses the following variables:

* **ymlepsq** - we transformed the existing two variables, **pctymle** (percent young male) and the **density** (people per square mile) to create this new variable which is the percent of young male per square mile in the population. 

* **polpc** - The second variable used in the regression model is the **polpc**, which is the police per capita

* **taxpc** - The third variable we used in the model is the tax revenue per capita

We selected these three independent variables to explain the criminal activity, with the hypothesis that there is a significant relationship between the percent of young males per capita in the population, police presence, tax revenue and the crime rate.

According to the results below, the model has achieved a good R-squared, about 68.7% of the crime rate is explained by the young males per square mile, the police per capita and tax revenue independent variables in the linear model.  The coefficients show that they are statistically significant.  Taking a look at the residuals which tells us the distance from the line at 0 is how bad the prediction was for that value.  We see some values which are greater than 0.02, which are some outlier values from the predicted values.  The maximum residual we observed was 0.04.

```{r, echo=FALSE}
filterC$ymlepsq = filterC$density * filterC$pctymle
model3 = lm(crmrte ~ ymlepsq + polpc + taxpc, data = filterC)
plot(model3$fitted.values, model3$residuals, main="Residuals vs Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
summary(model3)$r.square
stargazer(model3, type = "text",out = "model3.txt")
AIC(model3)
```

\  

**MODEL 2A WITH EXPLANATORY VARIABLE AND COVARIATES:**

The second model finalized uses the variables from model 1 and we added the following covariate:
* *prbarr* - probability of arrest

According to the results below, the model has achieved a good R-squared, about 73.0% of the crime rate is explained by the young males per square mile, the police per capita and tax revenue, and the covariate probability of arrest.  The coefficients show that they are statistically significant.  Taking a look at the residuals, we see some values which are greater than 0.02, which are some outlier values from the predicted values.  The maximum residual we observed was 0.035.

```{r, echo=FALSE}
model4 = lm(crmrte ~ ymlepsq + polpc + taxpc + prbarr, data = filterC)
plot(model4$fitted.values, model4$residuals, main="Residuals vs Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
summary(model4)$r.square
stargazer(model4, type = "text",out = "model4.txt")
AIC(model4)
```

\  

**MODEL 2B WITH EXPLANATORY VARIABLE AND COVARIATES:**
We added the probability of conviction in the model to understand how the results change or improve.  The results show that the R-squared doesn’t improve much and adding the probability of conviction does not yield a statistically significant coefficient either. 

According to the results below, the model R-squared remains steady as Model 2A, about 73.0% of the crime rate is explained by the young males per square mile, the police per capita and tax revenue, and the covariate probability of arrest and conviction.  Taking a look at the residuals, we see some values which are greater than 0.02, which are some outlier values from the predicted values.  The maximum residual we observed was 0.034.

```{r, echo=FALSE}
model5 = lm(crmrte ~ ymlepsq + polpc + taxpc + prbarr + prbconv, data = filterC)
plot(model5$fitted.values, model5$residuals, main="Residuals vs Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
summary(model5)$r.square
stargazer(model5, type = "text",out = "model5.txt")
AIC(model5)
```

\  

**MODEL 3 WITH EXPLANATORY VARIABLE AND ALL COVARIATES:**
In this model, we will look at the effect that adding all covariates have on the model.  In particular, we are adding on the covariates *prbpris* and *totalwg* to the previous model.  The variable *prbpris*, similar to *prbconv*, has little to no effect on the model.  The variable *totalwg* is a transformed variable that adds up all of the weekly wages in the county in order to understand how the income or affluency of the county affects crime rate. 

According to the information below, the R-squared value is very high at 74.13%, meaning that approximately 74% of crime rate can be attributed to young males per square mile, police per capita, tax revenue per capita, probability of arrests, probability of convictions, probability of prison sentence, and total wages.  This value is similar to the R-squared values from Model 2, which demonstrates that Model 2 exemplifies the best determinants of crime.  The minimal change in R-squared shows the robustness of the model. 

```{r, echo=FALSE}
filterC$totalwg = filterC$wcon + filterC$wtuc + filterC$wtrd + filterC$wfir + filterC$wser + filterC$wmfg + filterC$wfed + filterC$wsta + filterC$wloc
model6 = lm(crmrte ~ ymlepsq + polpc + taxpc + prbarr + prbconv + prbpris + totalwg, data = filterC)
plot(model6$fitted.values, model6$residuals, main="Residuals vs Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
summary(model6)$r.square
stargazer(model6, type = "text",out = "model6.txt")
AIC(model6)
```

We can then take a closer look at this model by looking at how the model changes from suburban and urban areas. This will help us understand if there is a difference in crime rate between the two regions.  To determine if an area is considered urban or suburban, we took a look at the urban variable.  If the value of the urban variable was 1, we considered that county urban.  If the urban variable was 0, we considered that county suburban. 

We used a two-sided independent t-test to compare the mean crime rate between urban and suburban areas.  The null hypothesis would be that the average crime rate in urban areas is equal to the average crime rate in suburban areas, and the alternative hypothesis is that these average crime rates are not equal.  Since we get a p-value of less than 0.01, we can reject the null hypothesis with high statistical significance and large practical significance.    

```{r, echo=FALSE}
filterUrban = filterC[filterC$urban == 1,]
filterSuburban = filterC[filterC$urban == 0,]

t.test(filterSuburban$crmrte, filterUrban$crmrte, alternative="two.sided", mu=0, paired=FALSE, var.equal=FALSE, conf.level = 0.95)
cohen.d(filterUrban$crmrte, filterSuburban$crmrte)
```

\  

**MODEL 3A WITH EXPLANATORY VARIABLE AND COVARIATES FOR SUBURBAN AREAS:**
For this model, we will look at how the explanatory variables and all the covariates affect crime rate specifically in suburban areas. This model is not as strong as our previous models because the R-squared value is 54.7%. 

```{r, echo=FALSE}
model7 = lm(crmrte ~ ymlepsq + polpc + taxpc + prbarr + prbconv + prbpris + totalwg, data = filterSuburban)
plot(model7$fitted.values, model7$residuals, main="Residuals vs Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
summary(model7)$r.square
stargazer(model7, type = "text",out = "model7.txt")
AIC(model7)
```

\  

**MODEL 3B WITH EXPLANATORY VARIABLE AND COVARIATES FOR URBAN AREAS:**
For this model, we will look at how all the variables affect crime rates in urban areas.  Because the sample size is only 8 counties, we do not have enough data to make conclusive statements of urban areas specifically.  However, these are areas where we should continue to monitor and collect data on because as found from the t-test, we can see that there is a higher crime rate in urban areas over suburban areas. 

```{r, echo=FALSE}
model8 = lm(crmrte ~ ymlepsq + polpc + taxpc + prbarr + prbconv + prbpris + totalwg, data = filterUrban)
plot(model8$fitted.values, model8$residuals, main="Residuals vs Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
summary(model8)$r.square
stargazer(model8, type = "text",out = "model8.txt")
AIC(model8)
```

\  

**Omitted Variables:** The variables that we have chosen to omit are *year*, *avgsen*, *west*, *central*, *pctmin80*, and *mix* because these variables have shown not to affect the crime rate. 

The variable *year* is not used because this remains constant at the value 1987 for all rows.  The variables *west* and *central* were not used because there were some counties that were not classified as either west or central so there was not an easy way to determine where these counties were regionally.  Because regions could not be determined for all counties, the *west* and *central* variables were omitted from the models. 

The variables *avgsen*, *mix*, and *pctmin80* are also not used because these variables had little to no effect on the crime rate.  This can be shown by looking at the R-squared value when correlating the variables with *crmrte*.  The R-squared values are very close to 0 indicating that these variables have little to no effect on the crime rate. 

```{r avgsen, echo=FALSE}
model_avgsen = lm(crmrte ~ avgsen, data = filterC)
summary(model_avgsen)$r.square
```
```{r west, echo=FALSE}
model_west = lm(crmrte ~ west, data = filterC)
summary(model_west)$r.square
```

```{r central, echo=FALSE}
model_central = lm(crmrte ~ central, data = filterC)
summary(model_central)$r.square
```

```{r pctmin80, echo=FALSE}
model_pctmin80 = lm(crmrte ~ pctmin80, data = filterC)
summary(model_pctmin80)$r.square
```

```{r mix, echo=FALSE}
model_mix = lm(crmrte ~ mix, data = filterC)
summary(model_mix)$r.square
```

Variable      | $R^2$
------------- | -------------
avgsen        | 0.02078
west          | 0.11246
central       | 0.02509
pctmin80      | 0.03286
mix           | 0.05403

Although mix is an omitted variable in the model specifications, we can still use this variable to analyze in more detail crime rate and total wages.

```{r, echo=FALSE}
plot(filterC$crmrte, filterC$totalwg, main = "Total Wage vs Crime Rate", xlab = "Crime Rate", ylab = "Total Wage")
model = lm(totalwg ~ crmrte, data = filterC)
abline(model)
```

```{r, echo=FALSE}
plot(filterC$totalwg, filterC$mix, main = "Mix vs Total Wage", xlab = "Total Wage", ylab = "Mix")
model = lm(mix ~ totalwg, data = filterC)
abline(model)
```

The first graph shows that as total wage increases crime rate also increases.  This means that there is more crime in more affluent areas.  This seems contrary to common conceptions, so we also looked at the relationship between total wage and face-to-face crimes.  From the graph of mix vs total wage, we can see that more face-to-face crimes, such as robberies, happen in less affluent areas.  So, although there is a higher crime rate in wealthier areas, there are more face-to-face crimes in lower income areas. 

\  

## 3. Conclusion

In conclusion, we can take our analysis to determine which policies would help decrease crime rate in North Carolina.  There are 3 major policies detailed below that we propose to help reduce crime rate.

**POLICY 1: INCREASE POLICE PRESENCE IN URBAN AREAS**
Because we are not able to determine a relationship between crime rate and west/central regions due to the limited information, we can provide a policy regarding the suburban and urban areas.  Based on our t-test findings, we can say that there is a higher crime rate in urban areas than in suburban areas with high statistical significance and large practical significance.  Although we cannot determine specifics regarding the crime rates in urban areas, the t-test helps to support a policy that increases police presence in urban areas.  It is also critical to collect more data on urban areas to continue to shape this policy to the specifications of urban areas. 

**POLICY 2: ADDED FOCUS ON WHITE COLLAR CRIMES IN AFFLUENT AREAS**
From our analysis, we can see that total wages in a county is a covariate variable when looking at crime rates.  As total wages increase, crime rates also increase.  This means that more affluent areas have higher crime rates.  Although higher crime rates tend to occur in wealthier areas,  less face-to-face crimes happen in these areas, indicating that these crimes are most likely white collar crimes.  Increasing the focus on white collar crimes in affluent areas will help to drive down crime rates. 

**POLICY 3: ALLOCATE RESOURCES BASED ON POPULATION DENSITY**
As backed by the analysis, one of the primary explanatory variables is density.  As population density increases, crime rates also increase.  Because this is the case, it would be best to allocate resources and police presence based on population density.  This would help to effectively utilize resources by taking out police forces in lower density areas and adding police forces in higher density areas. 

Taking into account these policy changes will help to reduce crime rate in North Carolina. 

\  

## 4. Appendix

```{r, echo=FALSE}
hist(filterC$ymlepsq, breaks = 20, main = "Histogram Representing YMLEPSQ", xlab = "Young Males Per Square Mile")
```

```{r, echo=FALSE}
plot(filterC$ymlepsq, filterC$crmrte, main = "Relationship between Crime and YMLEPSQ", xlab = "Young Males Per Square Miles", ylab = "Crime Rate")
model9 = lm(crmrte ~ ymlepsq, data = filterC)
abline(model9)
stargazer(model9, type = "text",out = "model9.txt")
```
